---
title: "Leaflet RUM Session"
author: "Reka Solymosi"
date: "2/2/2017"
output: html_document
---

#Intro

So this is a super brief intro into some of the cool things you can do with leaflet. There are comprehensive tutorials available online, for example [here](https://rstudio.github.io/leaflet/).

You will need to have installed the following packages to follow along:

```{r, eval=FALSE}
install.packages("leaflet") #for mapping

install.packages("RColorBrewer") #for getting nice colours for your maps

install.packages("rgdal") #for getting data from TfGM API
install.packages("httr") #for getting data from TfGM API
install.packages("jsonlite") #for getting data from TfGM API
```


#Making a map

To make a map, just load the leaflet library:

```{r}
library(leaflet)
```

You then create a map with this simple bit of code:

```{r}
m <- leaflet() %>%
  addTiles()  
```

And just print it:
```{r}
m  
```

#Adding some content:

You might of course want to add some content to your map. 

#Adding points manuall:

You can add a point manually:

```{r}
m <- leaflet() %>%
  addTiles()  %>% 
  addMarkers(lng=-2.230899, lat=53.464987, popup="You are here")
m  
```

Or many points manually:

```{r}

latitudes = c(53.464987, 53.472726, 53.466649) 
longitudes = c(-2.230899, -2.245481, -2.243421) 
popups = c("You are here", "Here is another point", "Here is another point") 
df = data.frame(latitudes, longitudes, popups)      

m <- leaflet(data = df) %>%
  addTiles()  %>%  
  addMarkers(lng=~longitudes, lat=~latitudes, popup=~popups)
m  
```

#Change the basemap

You can change the background as well. You can find a list of different basemaps [here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html).

```{r}
m <- leaflet(data = df) %>%
  addProviderTiles("Stamen.Toner") %>% 
  addMarkers(lng=~longitudes, lat=~latitudes, popup=~popups)
m  
```


#Adding data from elsewhere

You will most likely want to add data to your map form external sources, rather than manually creating points. 

For example, I illustrate here with data from [Manchester Open Data](http://open.manchester.gov.uk/) about public toilets: 

```{r}
publicToilets <- read.csv("http://www.manchester.gov.uk/open/download/downloads/id/171/public_toilets.csv")
```

Often spatial data will not come with latitude/longitude format, but with easting and northing. Leaflet (as far as I know) prefers lat/long so we might have to convert from BNG to WGS84. 

First thing we might notice is that the coordinates are in Easting and Northing format, rather than Latitude/ Longitude: 

```{r}
publicToilets[,8:9]
```

There is a comprehensive step-by-step tutorial on converting coordinates [here](http://www.alex-singleton.com/R-Tutorial-Materials/7-converting-coordinates.pdf). I'll just briefly demo this here.

##Reprojecting coordinates

```{r}
#the library I'm using here is rgdal
library(rgdal)

#these are the variables for the coordinate system types
bng = "+init=epsg:27700"
latlong = "+init=epsg:4326"

#create coords
coords <- cbind(Easting = as.numeric(as.character(publicToilets$GeoX)),
                Northing = as.numeric(as.character(publicToilets$GeoY)))

# create a SpatialPointsDataFrame
publicToiletsSPDF <- SpatialPointsDataFrame(coords, data = publicToilets, proj4string = CRS(bng))

#reproject with spTransform
publicToiletsSPDF_latlng <- spTransform(publicToiletsSPDF, CRS(latlong))

#extract coords into a column
publicToiletsSPDF_latlng@data$lng <- publicToiletsSPDF_latlng@coords[,1]
publicToiletsSPDF_latlng@data$lat <- publicToiletsSPDF_latlng@coords[,2]

```

Now you should have a reprojected spatial points data frame with latitude and longitude, ready to be mapped: 

```{r}
m <- leaflet(data = publicToiletsSPDF_latlng@data) %>%
  addProviderTiles("Stamen.Toner")  %>%  
  addMarkers(lng=~lng, lat=~lat, popup=~LocationText)
m  
```

#Add more meaning to your markers

You can also make your markers tell you something about your data. 

Let's look at a different dataset. 

This one from TfGM (accessed through their API). To use their API, you will have to get your own key. You can then use the code below, by replacing "enter your key here" with your key. To get your own key you simply have to [register with TFGM on their site for developers](https://developer.tfgm.com/). 

```{r, eval=FALSE}
api_key <- "enter your key here"
```

```{r, echo=FALSE}
fileName <- "tfgmKey.txt"
api_key <-readChar(fileName, file.info(fileName)$size)
```

Anyway once you have a key, you can get some data about various different transporty things. For example, here we get some information about car parks:

```{r}
library(httr) #library I use for getting data from API
library(jsonlite) #library I use for parsing the data into a dataframe

#get the data
req <- GET("https://api.tfgm.com/odata/Carparks?$expand=Location&$top=500", 
  add_headers("Ocp-Apim-Subscription-Key" = api_key))
stop_for_status(req)

#parse the data
thing <- content(req, as='text')
thing2 <- fromJSON(thing)

#finally get coordinates to columns from the WKT
thing2$value$lon <- as.numeric(paste0("-",gsub(".*?([0-9]+[.][0-9]+).*", "\\1", thing2$value$Location$LocationSpatial$Geography$WellKnownText)))
thing2$value$lat <- as.numeric(gsub(".* ([-]*[0-9]+[.][0-9]+).*", "\\1", thing2$value$Location$LocationSpatial$Geography$WellKnownText))

```

Now we have some live data about car parks! Let's map this! 

##Marker size and colous

We can set the size of the dots to indicate how many free spaces are currently available. We can also colour by a factor, let's say by the 'state' of the carpark. 

```{r}
library(RColorBrewer) #library for getting nice colours for your maps

#set colour scheme:
pal <- colorFactor("Paired", thing2$value$State) 

#can build a more complex popup by ysing paste:
popupText <- paste0("Name:",
                    thing2$value$Description,
                    "<br>",
                    "State: ",
                    thing2$value$State,
                    "<br>",
                    "Capacity: ",
                    thing2$value$Capacity)

leaflet(thing2$value) %>% 
  addProviderTiles("Stamen.Toner") %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = ~Capacity/75, #am making the numbers smaller otherwise we get giant blobs
    fillColor = ~pal(State),
    popup = popupText,
    stroke = 0.01,
    color = "black",
    fillOpacity = 1,
    weight=1
  ) 
```

Might also want to add a legend. Can do this with the `addLegend()` function. 

```{r}
leaflet(thing2$value) %>% 
  addProviderTiles("Stamen.Toner") %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = ~Capacity/75, #am making the numbers smaller otherwise we get giant blobs
    fillColor = ~pal(State),
    popup = popupText,
    stroke = 0.01,
    color = "black",
    fillOpacity = 1,
    weight=1
  ) %>%
  addLegend("bottomright", pal = pal, values = ~State,
    title = "State of carpark",
    opacity = 1)
```


##Cluster the markers
```{r}
leaflet(thing2$value) %>% 
  addProviderTiles("Stamen.Toner") %>%
  addMarkers(
    lng = ~lon,
    lat = ~lat,
    popup = ~Description,
    clusterOptions = markerClusterOptions()
  ) 
```

#Can also map polygons

Can also easily produce thematic maps with leaflet, or play around with polygons.

You can import a shapefile you might already have on your computer. For example, I have this

```{r}
fixMyStreet <- readOGR(dsn = "/Users/reka/Desktop/Data", "allDataAtLsoa")
fixMyStreet <- spTransform(fixMyStreet, CRS("+proj=longlat +datum=WGS84"))

```

```{r}
#make a fancy popup
boroughs_popup <- paste0("<strong>LSOA: </strong>",
                         fixMyStreet@data$LSOA11NM,
                         "<br><strong>Number of incivility reports: </strong>",
                         fixMyStreet@data$n_inciv)
#create colour palette
# rcolourbrewer gives loads to choose from, here are some examples: 

#DivergingBrBG, PiYG, PRGn, PuOr, RdBu, RdGy, RdYlBu, RdYlGn, Spectral

#QualitativeAccent, Dark2, Paired, Pastel1, Pastel2, Set1, Set2, Set3

#SequentialBlues, BuGn, BuPu, GnBu, Greens, Greys, Oranges, OrRd, PuBu, PuBuGn, PuRd, Purples, RdPu, Reds, YlGn, YlGnBu, YlOrBr, YlOrRd


pal <- colorNumeric(
  palette = "YlOrRd",
  domain = fixMyStreet@data$n_inciv
)

#make map
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons( data = fixMyStreet,
    stroke = TRUE, weight= 1, fillOpacity = 0.8, 
    color = ~pal(n_inciv), 
    popup = boroughs_popup
  ) %>%
#and add a Legend as well
  addLegend(pal = pal, 
            values = fixMyStreet@data$n_inciv,
            title = "Number of incivility reports",
            labFormat = labelFormat(suffix = " incivs"),
            opacity = 0.8
  )

```

##Mapping from geojson file

Can also map polygons from geojson files from the web. For example, here is a map of population density (number of persons per hectare) at the Output Area Level in Manchester: 

```{r}

geoData <- readLines("https://data.cdrc.ac.uk/dataset/35c1fb9d-df77-4261-9861-14fc75d6f26a/resource/da9e9f72-4d11-46e9-b33a-7a5e24b83d0c/download/cdrc-2013-mid-year-total-population-estimates-geodata-pack-lsoa-manchester-e08000003.geojson", warn = FALSE) %>%
  paste(collapse = "\n") %>%
  fromJSON(simplifyVector = FALSE)

value <- sapply(geoData$features, function(feat) {
  feat$properties$value
})

pal <- colorQuantile("Greens", value)
# Add a properties$style list to each feature
geoData$features <- lapply(geoData$features, function(feat) {
  feat$properties$style <- list(
    fillColor = pal(
      feat$properties$value)
    #)
  )
  feat
})

# Add the now-styled GeoJSON object to the map
leaflet() %>% setView(lng = -2.230899, lat=53.464987, zoom = 10) %>%
  addTiles() %>%
  addGeoJSON(geoData, weight=1, fillOpacity = 0.8) 

```

#Interactivity with Shiny

It is also possible to make the leaflet maps interactive, by integrating them with Shiny applications. The final dataset for this to experiment with comes from AccStats data for 2015, which can be accessed using the Transport for London API:

```{r, echo=FALSE}
fileName <- "tflApi.txt"
queryString <-readChar(fileName, file.info(fileName)$size)
```

```{r, eval=FALSE}
l = readLines(queryString, encoding="UTF-8", warn=FALSE)

d = fromJSON(l)

accidents <- data.frame(lapply(as.data.frame(d), as.character), stringsAsFactors=FALSE)

#also make sure data is in date format
accidents$date2 <- as.Date(accidents$date, "%Y-%m-%d")

```

NOTE: if you're going to be using this code, you will have to make a developer account with TfL to generate your own `app_id` and `app_key` for a query string. 
OK now with this data we can create an app for looking at the number of accidents with different severity, and adjust the date range (within the year 2015) as we like. 

First load shiny package. 
```{r}
library(shiny)
```

And then create an app, with the leaflet map inside: 

```{r, eval=FALSE}

ui <- fluidPage(
  titlePanel("Accidents in 2015"),

  sidebarLayout(
    sidebarPanel( 
      #date selector goes here 
      dateRangeInput("Date range", inputId = "date_range",
        start = "2015-01-01",
        end = "2015-12-31",
        format = "yyyy-mm-dd"), 
      uiOutput('severitySelector',selected = "Fatal")
      ),
    mainPanel(
      #leaflet output goes here
      leafletOutput("map", height = 800)

    )
  )
)

server <- function(input, output) {
  
  severityChoices <- sort(unique(as.character(accidents$severity)))
  
  #create the drop down menu with name country selector to put in placeholder in UI
  output$severitySelector <- renderUI({
	                          	selectInput("severitySelect", label = "Select severity",
					                    choices = as.list(severityChoices), selected = "Fatal")
  })
  #filter data based on dates
  dateFiltered <- reactive({
    thing <- accidents %>% filter(date2 %in% seq(input$date_range[1],     input$date_range[2], by = "day") & severity %in% input$severitySelect)
   
  })
  #reactive map
  output$map <- renderLeaflet({
    leaflet(accidents) %>%  
      addProviderTiles("CartoDB.Positron") %>%
      fitBounds(~min(lon), ~min(lat), ~max(lon), ~max(lat)) %>%
      addLegend(position = "bottomleft", colors = c("#b10026", "#fd8d3c", "#ffeda0"),
        labels = c("Fatal", "Serious", "Slight"), opacity = 1, title = "Severity")
    })
  
  observe({
    pal <- colorFactor(c("#b10026", "#fd8d3c", "#ffeda0"), domain = c("Fatal", "Serious", "Slight"), ordered = TRUE)
    leafletProxy("map", data = dateFiltered()) %>% clearMarkerClusters() %>%
      addCircleMarkers(~lon, ~lat,
        color = "#636363", stroke = TRUE, weight = 1,
        fillColor = ~pal(severity), fillOpacity = 0.8,
        radius = 5,
        popup = ~location, 
        clusterOptions = markerClusterOptions())
  })
  
  
}


shinyApp(ui = ui, server = server)

```

